---
title: "Misspecification of drift"
author: "Axel Szetu"
date: "2024-06-06"
output: html_document
---

The following notebook is written as an extension to the notebook AMOCestimation released as supplementary material to Ditlevsen et Al 2023.
THe purpose is to test the performance of the likelihood-based estimator of tipping time in a scenario where the drift is misspecified.
To this end, we will simulate realizations from a model with double-well potential and use the likelihood of 3'rd degree polynomial model to estimate the tipping time and simulate the distribution of the estimator.

We define the parameters of the simulation
```{r}
beta_4 <- 1
beta_2 <- 1
lambda_0 <- 1
lambda_c <- -0.4
t <- seq(0, 300, by = 1/12)
t_0 <- 100
tau <- 200
t_end <- t_0 + tau
h <- c(rep(0, 1200), seq(from = 0, to = 1, length.out = 2401))
lambda_t <- (1-h)*lambda_0 + h*lambda_c
x_0 <- 1.3
dt <- 1/12
nloop <- 20
sigma <- 0.5
```

We define a function to simulate a path given a list of parameters
```{r}
make_path <- function(x_0 = 1.3, betas = c(1,1), t_0 = 100, t_end = 300, lambda_0 = 1, lambda_c = -0.4, dt = 1/12, nloop = 2, sigma = 0.1){
  #browser()
  delta_t <- dt/nloop
  n_0 <- t_0/dt*nloop
  n_r <- (t_end-t_0)/dt*nloop
  n_total <- n_0 + n_r + 1
  t <- seq(from = 0, to = t_end, length.out = n_total)
  h <- c(rep(0, n_0), seq(from = 0, to = 1, length.out = n_r))
  lambda_t <- (1-h)*lambda_0 + h*lambda_c
  beta_4 <- betas[1]
  beta_2 <- betas[2]
  drift <- function(x, lambda){
    out <- -(beta_4*x^3) + beta_2*x + lambda
    out
  }
  X <- numeric(length = n_total)
  X[1] <- x_0
  dW <- rnorm(n = n_total, mean = 0, sd = 1)
  for (i in (2:n_total)){
    X[i] <- X[i-1] + drift(X[i-1],lambda_t[i-1])*delta_t + dW[i-1]*sigma*sqrt(delta_t)
  }
  out <- data.frame(t, X)
  out <- out[1 + nloop*(0:floor(n_total/nloop)),]
  out
}
test_path <- make_path()
plot(y = test_path$X, x = test_path$t)
```

We estimate parameters of OU
```{r}
data_OU <- test_path[test_path$t <= t_0, "X"]
temp   = estimate.OU(data = data_OU, delta = 1/12)
alpha0 = unname(coef(temp)["alpha"])
mu0    = unname(coef(temp)["mu"])
sigma2 = unname(coef(temp)["sigma2"])
```

We estimate tipping time
```{r}
data_ramp <- test_path[test_path$t > t_0, "X"]
temp = estimate.tipping(data = data_ramp, delta = 1/12, initial.values = c(200,2),
                        alpha0 = alpha0, mu0 = mu0, sigma20 = sigma2, 
                        pen = 0.004)
tau     = unname(temp$par[1])
a       = unname(temp$par[2])
m       = mu0 - alpha0/(2 * a)
lambda0 = -alpha0^2/(4 * a)
tc      = tau + t0
```
Collecting estimates
```{r}
## Collect estimates
round(c(t0 = t0, alpha0 = alpha0, mu0 = mu0, sigma2 = sigma2, tau = tau, 
        a = a, m = m, lambda0 = lambda0, tc = tc),2)
```


