---
title: "ROC curves for bifurcation detection, for presentation"
output: html_notebook
---
```{r}
library(ggplot2)
```


#ROC-curves for bifurcation detection
The purpose of this notebook is to
1) Develop a scheme for simulating the dynamics of a double-well potential with additive(for now) noise.
2) Compute hypothesis tests only using data from a restricted time interval.
3) Compute (and plot) ROC curves for different significance threshholds, control parameter speed, observation frequency and variance.

## Simulation of paths
```{r}
make_path <- function(x0, beta4 = 1, beta2 = 1, sigma = 1, tmax = 2, deltat = 0.01, lambda0 = 0.5, lambda_end){
  t <- seq(from = 0, to = tmax, by = deltat)
  x <- numeric(length = length(t))
  x[1] <- x0
  dW <- rnorm(n = length(t) - 1)
  lambda <- seq(from = lambda0, to = lambda_end, along.with = t)
  drift <- function(x, lambda){
    -beta4*x^3 + beta2*x + lambda
  }
  for (i in (2:length(t))){
    x[i] <- x[i-1] + drift(x = x[i-1], lambda = lambda[i-1])*deltat + dW[i]*sqrt(deltat)*sigma
  }
  data.frame(t,x)
}
```
We test that the function works and make plots for different speeds of lambda change:
```{r}
test_path <- make_path(x0 = 1, sigma = 0.05, tmax = 10, deltat = 0.01, lambda0 = 0, lambda_end = -1)
plot(test_path$x)
test_path2 <- make_path(x0 = 1, sigma = 0.05, tmax = 10, deltat = 0.01, lambda0 = 0, lambda_end = -0.4)
plot(test_path2$x)
test_path3 <- make_path(x0 = 1, sigma = 0.05, tmax = 10, deltat = 0.01, lambda0 = 0, lambda_end = -0.1)
plot(test_path3$x)
test_path4 <- make_path(x0 = 1, sigma = 0.05, tmax = 10, deltat = 0.01, lambda0 = 0, lambda_end = 0)
plot(test_path4$x)
```
In the first picture, it is clear that bifurcation has occured.
In the 2nd picture, we have just exacly reached the critical value.
In the 3rd picture, lambda has moved, but only 1/4 of the way to the critical value.
In the 4th picture lambda is constant.
The challenge is distinguishing between cases 3 and 4.

We desire a controlled experiment where we know the exact time of bifurcation.
Thus, we need to do some setup computation to establish the critical value of the control parameter.
By considering the derivative of the drift we find that it has two roots.
Superimposing the drift with the derivative of the drift shows that the larger root of the derivative is the value of x where the saddle-node bifurcation occurs. We can evaluate $F(x,\lambda)$ in $x = \frac{1}{sqrt{3}}$ and solve the resulting equation for lambda.
In the case where $\beta_4 = \beta_2 = 1$, we get $\lambda_c = \frac{2}{3 sqrt{3}}$.
See the following animation for a visualization:
https://www.geogebra.org/calculator/p9mqsm98

##Estimation and hypothesis testing
The following code implements a simple t-test that takes the residuals of a sample path as compared to a null-hypothesis and either rejects, og non-rejects the hypothesis.
```{r}
hypothesis_tester <- function(data, beta4 = 1, beta2 = 1, classifier = z_test){
  x <- data$x
  t <- data$t
  deltat <- t[2] - t[1]
  drift <- function(x){
    -beta4*x^3 + beta2*x
  }
  x_predicted <- numeric(length = length(x))
  x_predicted[1] <- x[1]
  for (i in (2:length(x_predicted))){
    x_predicted[i] <- x[i-1] + drift(x[i-1])*deltat
  }
  residuals <- (x-x_predicted)[2:length(x)]
  out <- z_test(residuals)
  out
}
```

```{r}
z_test <- function(x, alpha = 0.4){
  rejected <- 0
  Z <- mean(x, na.rm = TRUE)/sd(x, na.rm = TRUE)
  if (pnorm(abs(Z)) > 1-alpha/2){
    rejected <- 1 #Reject = true => bifurcation detected
  }
  out <- c(rejected, Z, pnorm(abs(Z)))
}
```

We test that the function works:
```{r}
test_path <- make_path(x0 = 1, sigma = 0.05, tmax = 10, deltat = 0.01, lambda0 = 0, lambda_end = -1)
hypothesis_tester(data = test_path)

```
## Computing ROC curves
To compute and plot ROC curves we need a vector of parameters along with their associated false positive rates and true positive rates.
The plan is as follows:
1) Provide vector of thresholds
2) Compute false positive rates
3) Compute true positive rates
4) Make graph
5) Compute AOC

### Computing positive rates, true and false
We should probably start with points 2 and 3.
A single MC function should do both, given different inputs: true and false inputs.
In order to use the same sample paths for different thresholds (and different classifiers) we make a function for generating the data and one for computing positive rate, given a classifier.
```{r}
sample_paths_maker <- function(model, n){
  sample_paths <- vector(mode = "list", length = n)
  for (i in (1:length(sample_paths))){
    sample_paths[[i]] <- model()
  }
  sample_paths
}
```

```{r}
positive_rate_computer <- function(data, method){
  n <- length(data)
  positives <- 0
  for (i in (1:n)){
    positive <- method(data[[i]])[1]
    positives <- positives + positive
  }
  rate <- positives/n
  rate
}
```


####Older version that didn't accomodate using same dataset for each threshold and classifier.
```{r}
positive_rate_computer <- function(model, method, n = 100){
  #browser()
  positives <- 0
  for (i in (1:n)){
    test_path <- model()
    positive <- method(test_path)[1]
    positives <- positives + positive
  }
  rate <- positives/n
  rate
}
```
Let's test the above function
```{r}
#Defining the parameters for the rate computation
model_H1 <- function(){
  make_path(x0 = 1, sigma = 0.05, tmax = 10, deltat = 0.01, lambda0 = 0, lambda_end = -1)
}
method <- hypothesis_tester
```
```{r}
true_positive_rate <- positive_rate_computer(model = model_H1, method = method, n = 100)
```

### Computing positive rates for a vector of thresholds
We propose the following procedure:
Create a list of thresholds (This probably has to be hand-made for each method)
Create a list of "methods" that is a given method for the different thresholds
Apply the positive rate function to the list of methods.
Do the above again for true/false.
Combine vectors and plot/compute AOC.

We start by making a vector of thresholds:
```{r}
u <- seq(from = 0.7, to = 0.8, length.out = 11)
thresh <- qnorm(u)[2:(length(u)-1)]
```
Now me make a function creating a list of methods - one for each threshold.
```{r}
z_tester_maker <- function(thresh){
  z_test <- function(x, alpha){
    rejected <- 0
    Z <- mean(x, na.rm = TRUE)/sd(x, na.rm = TRUE)
    if (pnorm(abs(Z)) > 1-(1-alpha)/2){
      rejected <- 1 #Reject = true => bifurcation detected
    }
    out <- c(rejected, Z)
  }
  hypothesis_tester <- function(data, beta4 = 1, beta2 = 1, classifier = z_test){
    x <- data$x
    t <- data$t
    deltat <- t[2] - t[1]
    drift <- function(x){
      -beta4*x^3 + beta2*x
    }
    x_predicted <- numeric(length = length(x))
    x_predicted[1] <- x[1]
    for (i in (2:length(x_predicted))){
      x_predicted[i] <- x[i-1] + drift(x[i-1])*deltat
    }
    residuals <- (x-x_predicted)[2:length(x)]
    out <- z_test(x = residuals, alpha = thresh)
    out
  }
  hypothesis_tester
}
```

```{r}
z_testers <- lapply(X = thresh, FUN = z_tester_maker)
```
In order to compute rates for each threshold on the same set of paths we need to restructure slightly.
In the example computed, the rate computer generates the paths that it classifies.
We need the rate computer to take the set of paths as input.
The above issue has been resolved.

#### Computing rates for each threshold.
Now, we make a function for computing rates given a vector of functions.
```{r}
positive_rates_computer <- function(data, methods){
  n <- length(data)
  l <- length(methods)
  rates <- numeric(length = l)
  for (i in (1:l)){
    rate <- positive_rate_computer(data = data, method = methods[[i]])
    rates[i] <- rate
  }
  rates
}
```
```{r}
true_positive_rates <- positive_rates_computer(data = data, methods = z_testers)
```
```{r}
model_H0 <- function(){
  make_path(x0 = 1, sigma = 0.05, tmax = 10, deltat = 0.01, lambda0 = 0, lambda_end = 0)
}
data_H0 <- sample_paths_maker(model = model_H0, n = 100)
```
```{r}
false_positive_rates <- positive_rates_computer(data = data_H0, methods = z_testers)
```

####Computing AUC for pairs of positive rates.
We proceed by Riemann sum approximation.
```{r}
AUC_computer <- function(data){
  x <- data[[1]]
  y <- data[[2]]
  no_intermediate_points <- length(x) + 1
  intermediate_points <- numeric(length = no_intermediate_points)
  for (i in (2:(no_intermediate_points - 1))){
    intermediate_points[i] <- (x[i-1] + x[i])/2
  }
  intermediate_points[no_intermediate_points] <- 1
  interval_sizes <- diff(intermediate_points)
  columns <- interval_sizes*y
  AUC <- sum(columns)
  out <- list(AUC)
  #out <- list(interval_sizes, y, columns, AUC)
}
```

#### Tuning parameters to find a problem that is not too easy
Below we make a small script for tinkering with the parameters of the model.
The purpose is the find a set of parameters (including thresholds) that give a nondegenerate ROC curve.

```{r}
sigma_test <- 0.1
tmax <- 10
deltat <- 0.1
u <- seq(from = 0.01, to = 0.2, length.out = 11)
thresh <- qnorm(u)[2:(length(u)-1)]
z_testers <- lapply(X = u, FUN = z_tester_maker)
model_H0 <- function(){
  make_path(x0 = 1, sigma = sigma_test, tmax = tmax, deltat = deltat, lambda0 = 0, lambda_end = 0)
}
data_H0 <- sample_paths_maker(model = model_H0, n = 100)
model_H1 <- function(){
  make_path(x0 = 1, sigma =sigma_test, tmax = tmax, deltat = deltat, lambda0 = 0, lambda_end = -0.1)
}
data_H1 <- sample_paths_maker(model = model_H1, n = 100)
false_positive_rates <- positive_rates_computer(data = data_H0, methods = z_testers)
true_positive_rates <- positive_rates_computer(data = data_H1, methods = z_testers)
positive_rates <- data.frame(false_positive_rates, true_positive_rates)
positive_rates
AUC_score <- 1-AUC_computer(data = positive_rates)[[1]]
AUC_score
AOC_curve <- ggplot(data = positive_rates, mapping = aes(x = x_test, y = y_test)) + geom_line()
AOC_curve
```
In the chunk above, the parameters give a problem that yields a meaningful ROC-curve.

